Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
After textual feedback Gerko

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multiple imputation by chained equations (MICE) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
 rticles::sage_article:
  keep_tex: yes
---

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous but also tedious. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI yields unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].

**Move this after terminology??** Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$. Missing data in $y$ is 'imputed' (i.e., filled in) $M$ times. The imputed data ($y_{imp}$) is combined with the observed data ($y_{obs}$) to create $M$ completed datasets. On each completed dataset, the analysis of scientific interest (the 'substantive model', or 'complete data model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained. These $M$ values are combined into a single pooled estimate $\bar{Q}$. **ALSO: make clear that iteration happens within the first step, between incomplete and completed data.**

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

@@

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed. For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice]. The convergence properties of this MI algorithm are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.

**Include sub-questions?:** How can non-convergence be diagnosed? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?

@@

## Terminology

@@

# How may non-convergence be diagnosed?

The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13]. Without mixing, imputation chains do not intermingle nicely. Without stationarity, there is trending within imputation chains, see Figure \ref{fig:non-conv}. **Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable. The first plot is typical convergence of MICE, the second is pathological non-convergence because of a mis-specified imputation model. Each line is an imputation. In the first plot, the lines intermingle nicely and there is little trending. In the second, there is a lot of trending and the lines do not cross much. Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm. This shows the importance of reaching converged states in iterative imputation algorithms.**

Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, mixing and stationarity will be evaluated separately in this study. As recommended [e.g., @cowl96, p. 898], $AC$ and $\widehat{R}$ will be used.

@@

From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$ **remove or explain why**. $\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. **Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.

## In practice/Answer
**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence diagnostics for MICE algorithms, the following condition must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance). And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**

**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together. So are the non-convergence diagnostics. Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations. As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up in the typical convergence situation. The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$. Panel C displays the $AC$ as computed with the R function stats::acf(). When we look at this panel, we must conclude that something weird is going on. The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on stats::acf(), we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms. There is a correction factor for a mathematical shortcut that works in the limit. According to... the function works when the number of iterations > 50. The default number of iterations in mice, however, is 5. Therefore, we compute $AC$ manually, see panel D. The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**

```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
load("./images/diagnostics_plot.Rdata")
diagnostics_plot
```

@@

Convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$). We expect that a completely converged state will not be reached.

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2]. That is, there is no baseline measure available to evaluate performance against.
<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$ **explain that the randomness induced by the MICE algorithm would take care of this??**. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic. **Added later: High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains. But this doesn't seem the case in practice??**

@@

The two challenges of convergence are stationarity and mixing. Mixing and stationarity can be inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances). As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient). A user-specified scalar summary, however, is not universal to all complete data problems. Therefore, as inspired by [@mack03], we will propose a $\theta$ that summarizes the multivariate state space of the algorithm. Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.

@@

The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands (quantities of scientific interest; $Q$s). It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased (and confidence valid) estimate of $Q$, the algorithm is sufficiently converged. For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research. The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates (**all variables or just for Y??**). Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$. Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**). **remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**

@@

Upon convergence, $\widehat{R}=1$ and $AC=0$, but in MCMC algorithms these are very unlikely (because it's a distribution, not a point to converge to). In practice, thresholds are used to diagnose non-convergence. Usually, non-convergence is diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether $AC$ is significantly different from zero. Performance measures are the same as above: unbiased, confidence valid estimates.

**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is not observed. But what about the thresholds for practice? $\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s. Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again. If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed. The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures. But it seems OK. $\widehat{R}$ < 1.01 is too stringent compared to the performance measures. And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).

@@

```
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
 for (number of simulation runs from 1 to 1000)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. compute convergence diagnostics
   5. perform analysis of scientific interest
   6. pool results across imputations
   7. compute performance measures
 8. aggregate outcomes across simulation runs 
9. combine outcomes of all missingness proportions
```
The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$), and 2) varying the missingness proportions ($p_{miss} =.05,.25,.50,.75,.95$). The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints. **As the number of iterations goes up, the imputation chains will become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem. The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions then cause slower convergence. However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) point will be reached.

The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). Let

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} 
In this study, sampling variance is not of interest. Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd]. A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`. Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem. Let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $i = 1, 2,..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.

We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely... solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are! Namely, the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$)**):

@@

The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`. (**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). This study only considers only an MCAR missingness mechanism. Therefore, results may not be extrapolated to other missing data problems. Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.

Missing datapoints in $y$ are imputed with the function `mice::mice()`. All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`). The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$). Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$). From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**). We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration. For each $\theta$ we estimate the corresponding scientific estimand $Q$.

>>>>

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Iterative imputation by chained equations (MICE) is a widely used tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence of iterative impuation algorithms."
keywords: MICE; convergence 
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
 rticles::sage_article:
  keep_tex: yes
---

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous but also **tedious**. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI has proven to be a powerful technique to yield unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].

**Move this after terminology??** Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$. Missing data in $y$ is 'imputed' (i.e., filled in) $M$ times. The imputed data ($y_{imp}$) is combined with the observed data ($y_{obs}$) to create $M$ completed datasets. On each completed dataset, the analysis of scientific interest is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained. These $M$ values are combined into a single pooled estimate $\bar{Q}$.

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on how to evaluate convergence of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

@@

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed. For reasons of brevity, we only focus on the MI algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. The convergence properties of the MICE algorithm are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.

**Include sub-questions?:** How can non-convergence be diagnosed? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? **How severe is it when the algorithm has not converged? And what are guidelines for practice?**

@@

## Terminology/Notation

@@

# Identifying non-convergence

There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13]. Without mixing, imputation chains do not intermingle nicely, **indicating that ...**. Without stationarity, there is trending within imputation chains, which implies that further iterations would yiled a different set of imputations. In Figure \ref{fig:non-conv}... **Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable. The first plot is typical convergence of MICE, the second is pathological non-convergence because of a mis-specified imputation model. Each line is an imputation. In the first plot, the chains intermingle nicely and there is little to no trending. In the second plot, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm. This shows the importance of reaching converged states in iterative imputation algorithms.**

Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, mixing and stationarity will be evaluated separately in this study. As recommended [e.g., @cowl96, p. 898], $AC$ and $\widehat{R}$ will be used **to diagnose non-convergence**.

@@

From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$ **remove or explain why, or leave in**. $\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. **Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of our scope.

## In practice/Answer
**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, the following condition must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance). And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**

**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together---as are the non-convergence diagnostics. Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations. As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up **in the typical convergence situation**. The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$. Panel C displays the $AC$ as computed with the R function `stats::acf()`. When we look at this panel, we conclude something weird. The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on stats::acf() [@R], we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms. There is a correction factor for a mathematical shortcut that works in the limit. According to @box15, the function works when the number of iterations > 50. The default number of iterations in iterative imputation, however, is often lower. Therefore, we compute $AC$ manually, see panel D. The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**

```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence. **ADD PANEL LABELS**"}
load("./images/diagnostics_plot.Rdata")
diagnostics_plot
```

@@

For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$). **We expect that a completely converged state will not be reached.**

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2]. **That is, there is no baseline measure available to evaluate performance against.**
<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In e.g. `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$ **explain that the randomness induced by the MICE algorithm would take care of this??**. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic. **Added later: High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains. But this doesn't seem the case in practice??**

@@

The two challenges of convergence are stationarity and mixing. Mixing and stationarity can be inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances). As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient). A user-specified scalar summary, however, is not universal to all complete data problems. Therefore, as inspired by [@mack03], we will propose a $\theta$ that summarizes the multivariate state space of the algorithm. Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets. **Explain model-dependent...**

@@

The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands, i.e. quantities of scientific interest $Q$. It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased and confidence valid estimate of $Q$, the algorithm is sufficiently converged. For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research. The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates (**all variables or just for Y??**). Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$. Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**). **remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**

@@

Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because **it's** convergence to a distribution. In practice,  non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether $AC$ is significantly different from zero. Performance measures are the same as above: unbiased, confidence valid estimates.

**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is never observed. But what about the thresholds for practice? $\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s. Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again. If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed. The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures. But it seems OK. $\widehat{R}$ < 1.01 is too stringent compared to the performance measures. And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).

@@

```
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes across simulation runs

```
The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$), and 2) varying the missingness proportions ($p_{miss} =.05,.25,.50,.75,.95$). The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints. **As the number of iterations goes up, the imputation chains will become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem. The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions then cause slower convergence. However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) point will be reached.

The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). Let **what?**

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} 
In this study, sampling variance is not of interest. Therefore, a single complete set may serve as comparative truth in all simulation runs [@vink14]. A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem. Let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $i = 1, 2,..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.

**We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$). We solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$**

@@

The complete data is *amputed* once for each simulation repetition with function `mice::ampute()`.

(**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data: The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). **

Proper performance of the convergence diagnostics under at least missing completely at random (MCAR) missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. However, results may not be extrapolated to other missingness mechanisms. **Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.**

Missing datapoints in $y$ are imputed with the `mice` package [@mice]. All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`). The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$). Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$). From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**). We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration. For each $\theta$ we estimate the corresponding scientific estimand $Q$.
